Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-06-20 paper.tex
upload

\begin{abstract}
  We present \emph{Decoupled Fallback}, a work-stealing method that enables single-pass \emph{Chained Scans} to run on hardware without \emph{forward-progress guarantees} (FPG) while avoiding \emph{starvation}. Additionally, we introduce a tile state representation for \emph{Chained Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation. On FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled Fallback} achieves near-\emph{Memcpy} speeds for inclusive prefix sum in WebGPU native and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach. We demonstrate that work-stealing is particularly well suited to \emph{Chained Scan}. By imposing blocking rates of up to 50\% to simulate unfair schedulers, we show that \emph{Decoupled Fallback} outperforms \emph{Reduce-then-Scan} even under extreme contention.
\end{abstract}

@@

\subsubsection{Inter-Workgroup Barriers}
Parallel programs in general require synchronization between multiple threads (those that don't are considered embarrassingly parallel). A standard technique for synchronization is \emph{barriers.} GPUs provide a limited palette of barrier primitives. At the coarsest granularity, \emph{pipeline barriers} fully synchronize both control and memory between multiple kernel launches. However, pipeline barriers are extremely expensive and efficient algorithms strictly minimize their use. At a much finer granularity, \emph{workgroup barriers} provide reasonably efficient synchronization between the threads in a workgroup. However, synchronization between workgroups is more limited. Standard GPU programming environments provide no control barrier between workgroups.\footnote{CUDA recently introduced \emph{Thread Block Cluster} synchronization~\cite{NvidiaCudaGuide}, but it is almost certainly designed to operate within the \emph{Persistent Thread}~\cite{gupta2012} paradigm, rather than serving as a barrier across an unbounded launch.} The availability of a memory-only barrier varies across GPU environments; it is present in Direct3D and Vulkan, but was missing in versions of Metal prior to 3.2. Because of its non-portability, it is also absent from the WebGPU specification. Even when available, the use of such memory barriers causes performance issues as they are usually implemented as a full cache flush.

@@

The viability of \emph{Decoupled Fallback} relies on the same cache locality properties as \emph{Decoupled Lookback}. Since no subgroup scheduler is perfectly fair, all hardware inherently has a natural fallback rate $f$\!, independent of FPG, due to false-positive fallback operations. As a result, memory bandwidth usage increases (and we expect throughput to decrease) by $O(fn)$ to $O((2 + f)n)$. However, just as serial ordering confines spine memory accesses within an $o$-width sliding window, scan input accesses are likewise constrained to an $ot$-width window, increasing their likelihood of residing in cache. As we later empirically demonstrate in Section~\ref{sec:Sim-Blocking}, this cache residency significantly reduces memory access costs, making fallback operations predominantly compute-bound when reducing a blocking tile. Since pure reduction has minimal overhead, throughput approaches $O(2n)$ on most devices.

@@

For a subgroup of size $s$, the Kogge-Stone subgroup network has a work complexity of $s \log_2 s$. Given an input of size $n$, the work complexity is:
\begin{equation}
  \begin{aligned}[b]
    \text{Work} & = \underbrace{\vphantom{\sum_{k=0}^{\lceil \log_s n \rceil}}n}_{\text{fanout}}
    +
    \underbrace{\sum_{k=1}^{\lceil \log_s n \rceil}}_{\text{iterations}}
    \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}\frac{n}{s^k}}_{\text{calls per iteration}}
    \cdot
    \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}s \log_2 s}_{\text{work per call}} \\
                & = n + s \log_2 s \cdot \sum_{k=1}^{\lceil \log_s n \rceil} \frac{n}{s^k}       \\
                & = O\left(n + n \log_2 s\right)                                                 \\
                & = O(n \log_2 s)                                                                \\
                & = O(n).
  \end{aligned}
\end{equation}
Since the subgroup size remains constant within each kernel launch, the final work complexity is $O(n)$. The Kogge-Stone scan has a depth of $\log_2 s$; the total depth of the workgroup-wide scan is:
\begin{equation}
  \begin{aligned}[b]
    \text{Depth} & = \underbrace{\lceil \log_s n \rceil}_{\text{loop iterations}}
    \cdot \underbrace{\log_2 s}_{\text{depth per iteration}}                           \\
                 & = \left\lceil \frac{\log_2 n}{\log_2 s} \right\rceil \cdot \log_2 s \\
                 & = \log_2 n + O(1).
  \end{aligned}
\end{equation}
Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes. However, while minimal depth is preserved regardless of the subgroup size, this result can be somewhat misleading, as the depth incurred during the subgroup portion of the scan requires no barrier, whereas the main loop incurs a workgroup-wide barrier at each iteration. Consequently, while the theoretical depth remains constant across subgroup sizes, smaller subgroups tend to perform worse in practice due to the increased relative cost of synchronization and the overhead of additional iterations in the main loop.

@@

      \If{$\textit{thread\_id} < \textit{step}$}{
        $\textit{temp} \gets \text{subgroupInclusiveScan}(\textit{x}[\textit{thread\_id} + \textit{top\_offset}])$\;
        $\textit{x}[\textit{thread\_id} + \textit{top\_offset}] \gets \textit{temp}$\;
        \If{$\text{\normalfont highestRankLane}(\textit{thread\_id})$}{
          $\textit{x}[(\textit{thread\_id} / \textit{S}) + \textit{step} + \textit{top\_offset}] \gets \textit{temp}$\;
        }
      }

@@

In particular, we use two or four 32-bit values, splitting the reduction value into 16-bit segments while reserving the two most significant bits of each 32-bit value to encode the tile state. Although this could also be accomplished with shared memory (and workgroup barriers), as an optimization, we use two or four threads from the first subgroup, leveraging subgroup \emph{ballot} to verify state consistency, and subgroup \emph{shuffle} to join and split reductions. With one thread per 16 bits and a minimum subgroup size of four in WebGPU, this method supports both 32-bit and 64-bit values, enabling 64-bit scans without 64-bit atomics or memory barriers.\footnote{Although native 64-bit values are not currently supported by the WebGPU specification, their future inclusion is anticipated.}

Fallbacks require a tile state representation and an atomic operation that can safely transition a tile to \emph{Ready} without overwriting a more advanced state. When an update attempt is made, a blocking tile may be in any of the three states due to either of two scenarios: (1) a false-positive fallback on a tile that transitions to \emph{Ready} or \emph{Inclusive} before fallback completion, or (2) another workgroup completing its fallback first, leaving the tile in \emph{Ready}. If the tile state is \emph{Inclusive} then the write must be skipped, especially because races in writes for the parts may leave the tile permanently in an unreadable state. Thus, a simple atomic store is not suitable. An atomic compare-and-swap would have the correct semantics, but WebGPU lacks this primitive (it does have \texttt{atomicCompareExchangeWeak}, which could possibly be made to work but makes reasoning about correctness and performance more difficult). Instead, we encode the tile state in the most significant bits of the atomic value and enforce a strict state ordering: \emph{Inclusive} $>$ \emph{Ready} $>$ \emph{Not Ready}. This allows an \texttt{atomicMaximum}, which successfully updates to a more advanced state but prevents a state regression, to correctly perform updates.

@@

\section{Evaluation}
\subsection{Experimental Setup}
We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$. We find a maximum spin count of 4 effectively balances responsiveness to blocking while minimizing false-positive fallbacks. As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4. Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB\@. For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum. Our test hardware, shown in Figure~\ref{tab:hardware}, spans a wide range of FPG and performance capabilities. We empirically find that Google Chrome's native implementation of WebGPU, Dawn, provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires native Android, necessitating the use of WGPU\@.\footnote{To keep the performance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
\begin{table*}
  \centering
  \begin{tabular}{l l c r r}
    \toprule
    Vendor & Device Name            & LOBE         & Occupancy & Subgroup Size \\
    \midrule
    Intel  & HD Graphics 620        & \textbf{Yes} & 9         & 8--32         \\
    ARM    & Mali-G78 MP20          & \textbf{No}  & 40        & 16            \\
    Apple  & M3     (10 Cores)      & \textbf{No}  & 80        & 32            \\
    Apple  & M1 Max (32 Cores)      & \textbf{No}  & 160       & 32            \\
    NVIDIA & GeForce RTX 2080 Super & \textbf{Yes} & 192       & 32            \\
    AMD    & Radeon RX 7900 XT      & \textbf{Yes} & 420       & 32--64        \\
    \bottomrule
  \end{tabular}
  \caption{Hardware Specifications of Testing Devices.\label{tab:hardware}}
\end{table*}

\subsection{Performance Comparison}
\subsubsection{Cross-API Performance Versus Contemporary Implementations}
We first compare \emph{Decoupled Fallback} (\emph{DF}) to two key benchmarks: \emph{Memcpy}\textdagger{A1} and the \emph{Decoupled Lookback} (\emph{DL}) implementation found in NVIDIA's CUB library, using the 2080 Super only. To isolate the impact of differing APIs, we implement and evaluate \emph{DF} in both CUDA and Dawn.\footnote{Due to WebGPU's 128~MB storage buffer limit, our maximum test size is limited to $2^{25}$ 32-bit elements.} \emph{Memcpy} serves as a baseline reference, as it shares the same $2n$ global memory movement as \emph{DF} but lacks compute overhead and is purely memory bandwidth bound.

@@

\subsubsection{Same-API Performance Versus Baseline Measures}
We next compare \emph{DF} against \emph{Reduce-then-Scan}\textdagger{A2} (\emph{RTS}), which serves as a baseline to highlight the performance improvements of our approach over the prior FPG-constrained technique. This evaluation is conducted using Dawn across the full range of tested devices.

\begin{table*}
  \centering
  \small
  \setlength{\tabcolsep}{1pt}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l r r r r r r}
    \toprule
    Metric & \!\!\!\!\!\!Intel HD620 & ARM Mali-G78 MP20 & Apple M3 & Apple M1 Max & Nvidia 2080 Super & AMD 7900 XT \\
    \midrule
    Avg \emph{Memcpy} (32-bit ele/sec $\cdot 10^9$)       & 1.48    & 2.48     & 10.82   & 37.42    & 50.95    & 73.68    \\
    Avg \emph{DF} (32-bit ele/sec $\cdot 10^9$)           & 1.51    & 2.19     & 10.87   & 36.85    & 51.25    & 83.27    \\
    Avg \emph{RTS} (32-bit ele/sec $\cdot 10^9$)          & 1.06    & 1.93     & 7.46    & 25.75    & 34.36    & 62.57    \\
    \emph{DF}/\emph{Memcpy}                              & 102.0\% & 88.4\%   & 100.5\% & 98.4\%   & 100.6\%  & 113.0\%  \\
    \emph{DF}/\emph{RTS}                                 & 142.6\% & 113.6\%  & 145.7\% & 143.1\%  & 149.1\%  & 133.1\%  \\
    Avg Spins per workgroup                              & 0.420   & 0.303    & 0.719   & 0.752    & 0.679    & 0.533    \\
    Avg Lookback Length per workgroup                    & 1.714   & 4.036    & 4.598   & 10.737   & 12.350   & 22.648   \\
    Avg Total Fallbacks Initiated                        & 446.183 & 245.404  & 449.961 & 1046.216 & 333.324  & 266.641  \\
    Avg Total Successful Insertions                      & 21.475  & 0.733    & 23.887  & 4.822    & 11.378   & 46.288   \\
    \bottomrule
  \end{tabular*}
  \caption{\emph{Decoupled Fallback} performance, Inclusive Prefix Sum, $\mathbf{2^{25}}$ 32-bit elements.\label{tab:results}}
\end{table*}

As shown in Table~\ref{tab:results}, \emph{DF} achieves at least a $\sim$40\% speedup over \emph{RTS} across all tested devices, closely approaching the theoretically expected $O(3n)/O(2n)$ 50\% speedup, except on the ARM Mali. \emph{DF} closely matches \emph{Memcpy} across most devices, deviating only on the AMD 7900 XT and ARM Mali, where \emph{DF} outperforms \emph{Memcpy} by 30.9\% and 12.8\%, respectively. This suggests that additional compute latency may mitigate paging effects by separating loads from stores. On the ARM Mali, \emph{DF} appears to underperform, achieving only a 36\% speedup instead of the expected 50\%. However, its nearly identical performance to \emph{Memcpy} indicates that the implementation is memory-bandwidth-bound, limiting further gains.

@@

@@

Furthermore, this data reveals two notable behaviors: (1) As demonstrated by the Apple M3, a scheduler can be fair yet still lack FPG, demonstrating that fairness and FPG are not necessarily correlated. (2) The ARM Mali-G78 MP20 exhibits the most unfair scheduling among all tested devices, with a fallback rate of 30.8\% relative to workgroups launched. However, the average number of spins per workgroup is only 1.477---far below our fallback trigger threshold of 4 spins. This suggests a bimodal distribution of waiting times, where workgroups either proceed immediately with minimal delay or experience prolonged stalls that trigger fallbacks.

\subsection{Simulated Blocking}%
\label{sec:Sim-Blocking}
To demonstrate the resilience of our technique to unfair schedulers, we simulate fallback rates significantly higher than those observed organically by intentionally omitting work tile reduction updates. Because these updates are skipped, the immediate successor workgroup reads \emph{Not Ready}, spinning and triggering a fallback. Starting with infrequent omissions---1 in 512 work tiles---we progressively increase the omission rate to near-continuous levels---1 in 2 work tiles---recording \emph{DF} statistics at each step.
\begin{figure*}
  \centering
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/fallbacksInitiated_plot.pdf}
    \caption{Total number of fallback operations triggered as simulated blocking rates increase. This statistic verifies that the simulation effectively induces fallbacks as intended.\label{fig:fallbacks_initiated}}
  \end{subfigure}\hfill
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/totalSpins_plot.pdf}
    \caption{Average number of waiting spins per workgroup as simulated blocking rates increase. Higher spin counts correlate with increased fallback operations, reinforcing the behavior observed in Figure~\ref{fig:fallbacks_initiated}.\label{fig:total_spins}}
  \end{subfigure}

As shown in Figure~\ref{fig:execution_time}, performance relative to a no-omission baseline remains above 85\% even at near 100\% fallback rates across all devices---except for the Intel HD620, where only a 25\% fallback rate is induced. Notably, this performance level remains well above the equivalent \emph{RTS} performance of 66\%; \emph{DF} is consistently faster than the previous standard for portable scans. While Figures~\ref{fig:fallbacks_initiated} and~\ref{fig:total_spins} confirm that fallbacks are being induced, this test primarily measures the impact of increasing fallback operations rather than true blocking behavior. This is because workgroups are still able to proceed after an omission. In true starvation, a blocking workgroup stalls before posting its reduction, preventing subsequent workgroups from making progress. Unlike our simulation, which allows workgroups to eventually continue, true starvation results in increased lookback lengths, as blocked workgroups cannot complete \emph{Chained Propagation} and update their flags to inclusive. Furthermore, starvation effectively leads to \emph{occupancy reduction}: execution resources remain available, but fewer subgroups are scheduled for execution, reducing opportunities for context switching to mitigate latency.
\section{Discussion}

@@

Our reexamination of \emph{Chained Scan} with dynamic assignment raises fundamental questions about the GPU progress models proposed by Sorensen et al.~\cite{sorensen2021}: Can all HSA algorithms be converted to OBE via dynamic assignment? Are there cases, excluding atomic memory errors, where LOBE guarantees termination but dynamic assignment with OBE does not? Is OBE all you need?

\subsection{Conclusion}
In this paper, we introduced \emph{Decoupled Fallback}, a single-pass \emph{Chained Scan} capable of achieving SOL performance across all progress models while remaining compatible with modern graphics APIs and hardware constraints. Our approach enables any workgroup to compute the reduction of any blocking tile, eliminating the risk of starvation. We reexamined the progress models proposed by Sorensen et al.~\cite{sorensen2021} in the context of dynamically assigned work, demonstrating that Chained Scan implementations require, at minimum, the OBE guarantee. Additionally, we highlighted the risks of running \emph{Decoupled Lookback} without FPG. Furthermore, we introduced a subgroup-accelerated yet subgroup-size-agnostic scan based on a generalized radix-$s$ Ladner-Fischer scan, along with a tile state representation that eliminates reliance on 64-bit atomics and memory barriers by distributing tile data across multiple values.

Our artifact is available in the following GitHub repository: https://anonymous.4open.science/status/Decoupled-Fallback-Paper-A33C. The repository contains per API requirements and building instructions.

>>>>

\begin{abstract}
  We present \emph{Decoupled Fallback}, a work-stealing method that enables single-pass \emph{Chained Scans} to run on hardware without \emph{forward-progress guarantees} (FPG) while avoiding \emph{starvation}. Additionally, we introduce a tile state representation for \emph{Chained Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation. On FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled Fallback} achieves near-\emph{Memcpy} speeds for inclusive prefix sum in WebGPU native and approaches the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach. We further demonstrate the resilience of \emph{Decoupled Fallback} against \emph{unfair} schedulers by simulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
\end{abstract}

@@

\subsubsection{Inter-Workgroup Barriers}
Parallel programs in general require synchronization between multiple threads (those that don't are considered embarrassingly parallel). A standard technique for synchronization is \emph{barriers.} GPUs provide a limited palette of barrier primitives. At the coarsest granularity, \emph{pipeline barriers} fully synchronize both control and memory between multiple kernel launches. However, pipeline barriers are extremely expensive and efficient algorithms strictly minimize their use. At a much finer granularity, \emph{workgroup barriers} provide reasonably efficient synchronization between the threads in a workgroup. However, synchronization between workgroups is more limited. Standard GPU programming environments provide no control barrier between workgroups.\footnote{CUDA recently introduced \emph{Thread Block Cluster} synchronization~\cite{NvidiaCudaGuide}, but it is almost certainly designed to operate within the \emph{Persistent Thread}~\cite{gupta2012} paradigm, rather than serving as a global barrier across an unbounded launch.} The availability of a memory-only barrier varies across GPU environments; it is present in Direct3D and Vulkan, but was missing in versions of Metal prior to 3.2. Because of its non-portability, it is also absent from the WebGPU specification. Even when available, the use of such memory barriers causes performance issues as they are usually implemented as a full cache flush.

@@

The viability of \emph{Decoupled Fallback} relies on the same cache locality properties as \emph{Decoupled Lookback}. Since no subgroup scheduler is perfectly fair, all hardware inherently has a natural fallback rate $f$\!, independent of FPG, due to false-positive fallback operations. As a result, memory bandwidth usage increases (and we expect throughput to decrease) by $O(fn)$ to $O((2 + f)n)$. However, just as serial ordering confines spine memory accesses within an $o$-width sliding window, scan input accesses are likewise constrained to an $ot$-width window, increasing their likelihood of residing in cache. This cache residency significantly reduces memory access costs, making fallback operations predominantly compute-bound when reducing a blocking tile. Since pure reduction has minimal overhead, throughput approaches $O(2n)$ on most devices.

@@

For a subgroup of size $s$, the Kogge-Stone subgroup network has a work complexity of $s \log_2 s$. Given an input of size $n$, the work complexity is:
\begin{equation}
  \begin{aligned}[b]
    \text{Work} & = \underbrace{\vphantom{\sum_{k=0}^{\lceil \log_s n \rceil}}n}_{\text{fanout}}
    +
    \underbrace{\sum_{k=1}^{\lceil \log_s n \rceil}}_{\text{iterations}}
    \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}\frac{n}{s^k}}_{\text{calls per iteration}}
    \cdot
    \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}s \log_2 s}_{\text{work per call}} \\
                & = n + s \log_2 s \cdot \sum_{k=1}^{\lceil \log_s n \rceil} \frac{n}{s^k}       \\
                & = O\left(n + n \log_2 s\right)                                                 \\
                & = O(n \log_2 s)                                                                \\
                & = O(n).
  \end{aligned}
\end{equation}
Since the subgroup size remains constant within each kernel launch, the final work complexity is $O(n)$. The Kogge-Stone scan has a depth of $\log_2 s$; the total depth of the workgroup-wide scan is:
\begin{equation}
  \begin{aligned}[b]
    \text{Depth} & = \underbrace{\lceil \log_s n \rceil}_{\text{loop iterations}}
    \cdot \underbrace{\log_2 s}_{\text{depth per iteration}}                           \\
                 & = \left\lceil \frac{\log_2 n}{\log_2 s} \right\rceil \cdot \log_2 s \\
                 & = \log_2 n + O(1).
  \end{aligned}
\end{equation}
Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes. However, while minimal depth is preserved regardless of the subgroup size, this result can be somewhat misleading, as the depth incurred during the subgroup portion of the scan requires no barrier, whereas the main loop incurs a barrier at each iteration. Consequently, while the theoretical depth remains constant across subgroup sizes, smaller subgroups tend to perform worse in practice due to the increased relative cost of synchronization and the overhead of additional iterations in the main loop.

@@

      \If{$\textit{thread\_id} < \textit{step}$}{
        $\textit{temp} \gets \text{subgroupKoggeStoneScan}(\textit{x}[\textit{thread\_id} + \textit{top\_offset}])$\;
        $\textit{x}[\textit{thread\_id} + \textit{top\_offset}] \gets \textit{temp}$\;
        \If{$\text{\normalfont highestRankLane}(\textit{thread\_id})$}{
          $\textit{x}[(\textit{thread\_id} / \textit{S}) + \textit{step} + \textit{top\_offset}] \gets \textit{temp}$\;
        }
      }

@@

In particular, we use two or four 32-bit values, splitting the reduction value into 16-bit segments while reserving the two most significant bits of each 32-bit value to encode the tile state. Although this could also be accomplished with shared memory (and workgroup barriers), as an optimization, we use two or four threads from the first subgroup, leveraging subgroup \emph{ballot} to verify state consistency, and subgroup \emph{shuffle} to join and split reductions. With one thread per 16 bits and a minimum subgroup size of four in WebGPU, this method supports both 32-bit and 64-bit values, enabling 64-bit scans without 64-bit atomics or memory barriers.

Fallbacks require a tile state representation and an atomic operation that can safely transition a tile to \emph{Ready} without overwriting a more advanced state. When an update attempt is made, a blocking tile may be in any of the three states due to either of two scenarios: (1) a false-positive fallback on a tile that transitions to \emph{Ready} or \emph{Inclusive} before fallback completion, or (2) another workgroup completing its fallback first, leaving the tile in \emph{Ready}. If the tile state is \emph{Inclusive} then the write must be skipped, especially because races in writes for the parts may leave the tile permanently in an unreadable state. Thus, a simple atomic store is not suitable. An atomic compare-and-swap would have the correct semantics, but WebGPU lacks this primitive (it does have \emph{atomicCompareExchangeWeak}, which could possibly be made to work but makes reasoning about correctness and performance more difficult). Instead, we encode the tile state in the most significant bits of the atomic value and enforce a strict state ordering: \emph{Inclusive} $>$ \emph{Ready} $>$ \emph{Not Ready}. This allows an \emph{atomicMaximum}, which successfully updates to a more advanced state but prevents a state regression, to correctly perform updates.

@@

\section{Evaluation}
\subsection{Experimental Setup}
We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$. We find a maximum spin count of 4 effectively balances responsiveness to blocking while minimizing false-positive fallbacks. As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4. Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB\@. For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum. Our test hardware, shown in Figure~\ref{tab:hardware}, spans a wide range of FPG and performance capabilities. We empirically find that Google Chrome's native implementation of WebGPU, Dawn, provides the best performance and use it for all testing.
\begin{table*}
  \centering
  \begin{tabular}{l l c r r}
    \toprule
    Vendor & Device Name            & FPG         & Occupancy & Subgroup Size \\
    \midrule
    Intel  & HD Graphics 620        & \textbf{Yes} & 9         & 8--32         \\
    ARM    & Mali-G78 MP20          & \textbf{No}  & 40        & 16            \\
    Apple  & M3     (10 Cores)      & \textbf{No}  & 80        & 32            \\
    Apple  & M1 Max (32 Cores)      & \textbf{No}  & 160       & 32            \\
    NVIDIA & GeForce RTX 2080 Super & \textbf{Yes} & 192       & 32            \\
    AMD    & Radeon RX 7900 XT      & \textbf{Yes} & 420       & 32--64        \\
    \bottomrule
  \end{tabular}
  \caption{Hardware Specifications of Testing Devices; FPG determinations are based on the data collected by Sorensen et. al.~\cite{sorensen2021}\label{tab:hardware}}
\end{table*}

\subsection{Performance Comparison}
\subsubsection{Cross-API Performance Versus Contemporary Implementations}
We first compare \emph{Decoupled Fallback} (\emph{DF}) to two key benchmarks: \emph{Memcpy} and the \emph{Decoupled Lookback} (\emph{DL}) implementation found in NVIDIA's CUB library, using the 2080 Super only. To isolate the impact of differing APIs, we implement and evaluate \emph{DF} in both CUDA and Dawn.\footnote{Due to WebGPU's 128~MB storage buffer limit, our maximum test size is limited to $2^{25}$ 32-bit elements.} \emph{Memcpy} serves as a baseline reference, as it shares the same $2n$ global memory movement as \emph{DF} but lacks compute overhead and is purely memory bandwidth bound.

@@

\begin{table*}
  \centering
  \small
  \setlength{\tabcolsep}{1pt}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l r r r r r r}
    \toprule
    Metric & \!\!\!\!\!\!Intel HD620 & ARM Mali-G78 MP20 & Apple M3 & Apple M1 Max & Nvidia 2080 Super & AMD 7900 XT \\
    \midrule
    Avg \emph{Memcpy} (32-bit ele/sec $\cdot 10^9$)      & 1.48    & 2.50     & 10.82   & 37.42    & 50.95    & 73.68    \\
    Avg \emph{DF} (32-bit ele/sec $\cdot 10^9$)          & 1.51    & 2.65     & 10.87   & 36.85    & 51.25    & 83.27    \\
    Avg \emph{RTS} (32-bit ele/sec $\cdot 10^9$)         & 1.06    & 1.94     & 7.46    & 25.75    & 34.36    & 62.57    \\
    \emph{DF}/\emph{Memcpy}                              & 102.0\% & 106.0\%  & 100.5\% & 98.4\%   & 100.6\%  & 113.0\%  \\
    \emph{DF}/\emph{RTS}                                 & 142.6\% & 136.6\%  & 145.7\% & 143.1\%  & 149.1\%  & 133.1\%  \\
    Avg Spins per workgroup                              & 0.420   & 2.52     & 0.719   & 0.752    & 0.679    & 0.533    \\
    Avg Lookback Length per workgroup                    & 1.714   & 3.52     & 4.598   & 10.737   & 12.350   & 22.648   \\
    Avg Total Fallbacks Initiated                        & 446.183 & 4739.91  & 449.961 & 1046.216 & 333.324  & 266.641  \\
    Avg Total Successful Insertions                      & 21.475  & 571.24   & 23.887  & 4.822    & 11.378   & 46.288   \\
    \bottomrule
  \end{tabular*}
  \caption{\emph{Decoupled Fallback} performance, Inclusive Prefix Sum, $\mathbf{2^{25}}$ 32-bit elements.\label{tab:results}}
\end{table*}

\subsubsection{Same-API Performance Versus Baseline Measures}
We next compare \emph{DF} against \emph{Reduce-then-Scan} (\emph{RTS}), which serves as a baseline to highlight the performance improvements of our approach over the prior FPG-constrained technique. This evaluation is conducted using Dawn across the full range of tested devices.

As shown in Table~\ref{tab:results}, \emph{DF} achieves at least a $\sim$40\% speedup over \emph{RTS} across all tested devices, closely approaching the theoretically expected $O(3n)/O(2n)$ 50\% speedup, except on the 7900XT and Mali-G78. This result initially appears confusing, given that \emph{DF} outperforms \emph{Memcpy} by 6\% and 13\%, respectively. However, we hypothesized and confirmed a secondary effect of \emph{Chained Scans}: they enforce a queue-like behavior among workgroups. In-flight workgroups with a lower relative rank traverse fewer tiles during their lookbacks, enabling them to exit earlier than higher-ranked workgroups. Because this ordering emerges naturally from differing amounts of computation rather than explicit synchronization, the exit sequence is \emph{k-sorted} or \emph{nearly ordered}, rather than strictly ordered. Scheduling fairness and message-passing visibility can further influence this sequence---for example, an unfair scheduler might allocate more resources to a higher-ranked workgroup, causing it to exit before a lower-ranked counterpart, or a higher-ranked workgroup might observe an updated flag sooner, enabling earlier completion. We confirmed this behavior experimentally by using atomics to record and print the order in which workgroups completed their lookback phase. Unlike a simple \emph{Memcpy} kernel, where all in-flight workgroups begin writing immediately after completing reads, this “ordered exit” pattern facilitates more efficient global memory accesses during the write phase of the scan, as adjacent workgroups are more likely to share memory pages, reducing premature page evictions. We hypothesize that the 7900XT and Mali-G78 have insufficient memory systems to efficiently accommodate all in-flight workgroups, making them particularly sensitive to this effect. To test this hypothesis, we augmented our \emph{Memcpy} kernel with a simplified lookback routine---excluding the scan itself---and observed a performance improvement of approximately 10\% on the 7900XT. Thus, the unexpectedly slow baseline \emph{Memcpy} performance on these devices explains why \emph{DF}, despite its strong overall performance, does not fully achieve the theoretical 50\% speedup over \emph{RTS}.

@@

\label{sec:Sim-Blocking}
To demonstrate the resilience of our technique to unfair schedulers, we simulate fallback rates significantly higher than those observed organically by intentionally omitting work tile reduction updates. Because these updates are skipped, the immediate successor workgroup reads \emph{Not Ready}, spinning and triggering a fallback. Starting with infrequent omissions---1 in 512 work tiles---we progressively increase the omission rate to near-continuous levels---1 in 2 work tiles---recording \emph{DF} statistics at each step.
\begin{figure*}
  \centering
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/fallbacksInitiated_plot.pdf}
    \caption{Total number of fallback operations triggered as simulated blocking rates increase. This statistic verifies that the simulation effectively induces fallbacks as intended.\label{fig:fallbacks_initiated}}
  \end{subfigure}\hfill
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/totalSpins_plot.pdf}
    \caption{Average number of waiting spins per workgroup as simulated blocking rates increase. Higher spin counts correlate with increased fallback operations, reinforcing the behavior observed in Figure~\ref{fig:fallbacks_initiated}.\label{fig:total_spins}}
  \end{subfigure}

@@

Furthermore, this data reveals two notable behaviors: (1) As demonstrated by the Apple M3, a scheduler can be fair yet still lack FPG, demonstrating that fairness and FPG are not necessarily correlated. (2) The ARM Mali-G78 MP20 exhibits the most unfair scheduling among all tested devices, with a fallback rate of 57.8\% relative to workgroups launched. However, the average number of spins per workgroup is only 2.52---below our fallback trigger threshold of 4 spins. This suggests a bimodal distribution of waiting times, where workgroups either proceed immediately with minimal delay or experience prolonged stalls that trigger fallbacks.

\subsection{Simulated Blocking}%
As shown in Figure~\ref{fig:execution_time}, performance relative to a no-omission baseline remains above 85\% even at near-100\% fallback rates across all devices---except for the Intel HD620, where only a 25\% fallback rate is induced. Notably, this performance level remains well above the equivalent \emph{RTS} performance of 66\%; \emph{DF} is consistently faster than the previous standard for portable scans. While Figures~\ref{fig:fallbacks_initiated} and~\ref{fig:total_spins} confirm that fallbacks are being induced, this test primarily measures the impact of increasing fallback operations rather than true blocking behavior. This is because workgroups are still able to proceed after an omission. In true starvation, a blocking workgroup stalls before posting its reduction, preventing subsequent workgroups from making progress. Unlike our simulation, which allows workgroups to eventually continue, true starvation results in increased lookback lengths, as blocked workgroups cannot complete \emph{Chained Propagation} or update their flags to inclusive. Furthermore, starvation effectively leads to \emph{occupancy reduction}: execution resources remain available, but fewer subgroups are scheduled for execution, reducing opportunities for context switching to mitigate latency. The slight downward trend in Figure~\ref{fig:lookback_length} shows that as workgroups get bogged down in fallbacks, lookback lengths decrease as dependency resolution reverts to tile-to-adjacent-tile propagation of readiness information. This emergent behavior caused by blocking can be viewed as a computationally enforced variant of \emph{StreamScan}. We connect this behavior to \emph{StreamScan} because we posit that its strictly enforced adjacent-tile dependency propagation yields perfectly ordered workgroup exits, which could further enhance global write efficiency, possibly explaining the slight uptick in performance seen in Figure~\ref{fig:execution_time} on the Mali-G78.

\section{Discussion}

@@

\subsection{Conclusion}
In this paper, we introduced \emph{Decoupled Fallback}, a single-pass \emph{Chained Scan} capable of achieving SOL performance across all progress models while remaining compatible with modern graphics APIs and hardware constraints. Our approach enables any workgroup to compute the reduction of any blocking tile, eliminating the risk of starvation. We highlighted the risks of running \emph{Decoupled Lookback} without FPG, and introduced a subgroup-accelerated yet subgroup-size-agnostic scan based on a generalized radix-$s$ Ladner-Fischer scan, along with a tile state representation that eliminates reliance on 64-bit atomics and memory barriers by distributing tile data across multiple values.

Our artifact is available in the following GitHub repository: https://github.com/b0nes164/Decoupled-Fallback-Paper. The repository contains per API requirements and building instructions.

\begin{acks}
We thank Elias Naur for his initial work on Scalar Fallback, Ken Russell for facilitating this colloboration, Duane Merrill for answering our questions about scan terminology, and Oliver Le for providing the initial testing on Apple devices.
\end{acks}
